The hydra and the rooster

### 2020-07-06 latex/chapter-skippable.tex
work on type classes for ordinal notations

@@

\begin{Coqsrc}
Coercion is_true: bool >-> Sortclass.

@@

>>>>

\subsection{A type class for ordinal notations}

Let $<$ be some relation on a given type $A$. In order to build an ordinal  notation 
system out of $(A,<)$, we require $<$ to be a strict, well-founded order, and that 
the reflexive closure of $<$ is a total order on $A$.

\index{Coq!Techniques!Type classes}

Using \coq's standard library and type classes, we propose the following definition.

\vspace{4pt}
\noindent\emph{From Module~\href{../src/html/hydras.Prelude.Ordinal_generic.html}{Prelude.Ordinal\_generic}}

\begin{Coqsrc}
Class OrdinalNotation {A:Type}{lt: relation A}(sto:StrictOrder lt)
      (compare : A -> A -> comparison) :=
  { compare_correct :
      forall alpha beta:A,
        CompareSpec (alpha=beta) (lt alpha beta) (lt beta alpha)
                                 (compare alpha beta);
    wf : well_founded lt}.  
\end{Coqsrc}

@@

\label{def: Finite-ord-type}
\begin{Coqsrc}
Coercion is_true: bool >-> Sortclass.

@@

In order to build an instance of \texttt{OrdinalNotation}, we define a comparison function, by delegation to standard library's  \texttt{Nat.compare}, and prove its correction.

\begin{Coqsrc}
Definition compare {n:nat} (alpha beta : t n) :=
  Nat.compare (proj1_sig alpha) (proj1_sig beta).

Lemma compare_correct {n} (alpha beta : t n) :
  CompareSpec (alpha = beta) (lt alpha beta) (lt beta alpha)
              (compare alpha beta).
\end{Coqsrc}

\begin{remark}
 The proof of \texttt{compare\_correct} uses a well-know pattern of \coq{}.
Let us consider  the following subgoal.

\begin{Coqanswer}
 1 subgoal (ID 110)

  n, x0 : nat
  i, i0 : x0 <? S n
  ============================
  exist (fun i1 : nat => i1 <=? n) x0 i =
  exist (fun i1 : nat => i1 <=? n) x0 i0
\end{Coqanswer}

Applying the tactic \texttt{f\_equal} generates a simpler subgoal.

\begin{Coqanswer}
1 subgoal (ID 112)

  n, x0 : nat
  i, i0 : x0 <? S n
  ============================
  i = i0
\end{Coqanswer}

We have now to prove that two proofs of \texttt{Nat.ltb x0 (S n)} are equal.

This is not obvious, but  a consequence of the following lemma of library 
\href{https://coq.inria.fr/distrib/current/stdlib/Coq.Logic.Eqdep_dec.html}{Coq.Logic.Eqdep\_dec}.

\index{Coq!Techniques!Unicity of equality proofs}

\begin{Coqanswer}
eq_proofs_unicity_on :
forall (A : Type) (x : A),
(forall y : A, x = y \/ x <> y) -> forall (y : A) (p1 p2 : x = y), p1 = p2
\end{Coqanswer}

Thus unicity of proofs of \texttt{Nat.ltb x0 (S n)}  comes from the decidability of
equality on type \texttt{bool}.

This is why we used the boolean function \texttt{Nat.ltb} instead of the inductive predicate \texttt{Nat.lt} in the definition of type \texttt{t $n$} (see page~\pageref{def: Finite-ord-type}).

For more information about this pattern, please lokk at the numerous mailing lists and 
FAQs on \coq{}).

\end{remark}

Applying lemmas of the libraries \texttt{Coq.Wellfounded.Inverse\_Image},
 \texttt{Coq.Wellfounded.Inclusion}, and \texttt{Coq.Arith.Wf\_nat}, we prove that our
relation \texttt{wf} is well founded.

\begin{Coqsrc}
Lemma lt_wf (n:nat) : well_founded (@lt n).
\end{Coqsrc}

Now we can build our instance of \texttt{OrdinalNotation}.

\begin{Coqsrc}
Global Instance sto n : StrictOrder (@lt n).

Global Instance FinOrd (n:nat) : OrdinalNotation (sto n) compare.
Proof.
  split.
  - apply compare_correct.
  - apply lt_wf.
Qed.
\end{Coqsrc}

Finally, it is interesting to show that the segment $[0,n[$ is a ``sub-segment'' of
$[0,n+1[$. Since types $t\;n$ are $t\;(n+1)$ are not convertible, we consider a ``cast'' 
function from $t\;n$ to $t\;(n+1)$, and prove this function is compatible with
comparison function of both types, and that it is a bijection from $t\;n$ to
the segment $[0,n-1[$ of $t\;(n+1)$.

In other words, \texttt{FinOrd $n$}  is a subnotation of \texttt{FinOrd $(n+1)$}.

\begin{Coqsrc}
Class  SubNotation {A:Type}
       {ltA: relation A}{stoA:StrictOrder ltA}
       {compareA : A -> A -> comparison}
       (notA : OrdinalNotation stoA compareA)
       {B:Type}{ltB: relation B}{stoB:StrictOrder ltB}
       {compareB : B -> B -> comparison}
       (notB : OrdinalNotation stoB compareB)
       (bound :  B)
       (cast : A -> B):=
  {
  commutation :forall a a' : A,  compareB (cast a) (cast a') =
                                 compareA a a';
  incl : forall a, ltB (cast a) bound;
  surj : forall b, ltB b bound -> exists a:A, cast a = b}.
\end{Coqsrc}

Coming back to finite ordinals, the conversion function from $t\;i$ to $t\;j$ is
defined with \texttt{Program}.

\begin{Coqsrc}
Program Definition cast {i j : nat}  (H: i < j) (alpha: t i) : t j :=
  alpha.

Lemma cast_compare_commute (i j  :nat)(H : i < j) :
  forall alpha beta, compare alpha beta = compare (cast  H alpha)
                                                  (cast  H beta).

\end{Coqsrc}

We are now able to build an instance of \texttt{SubNotation}

\begin{Coqsrc}
(** biggest element of t (S n) *)

Program Definition biggest_S (n:nat) : t (S n) :=n.
Next Obligation. 
  red;rewrite  Nat.ltb_lt; auto with arith.
Defined.

Instance F_incl n : SubNotation  (FinOrd n) (FinOrd (S n)) (biggest_S n)
                                 (cast (Nat.lt_succ_diag_r n)).
\end{Coqsrc}
